---
title: "Final Project"
author: "Hirsch-Santagata, Pulliam"
format: html
editor: visual
---

```{r}
library(tidyverse) # for data wrangling and plotting
library(car) # for Anova function
library(lme4)
library(broom) # for model residuals extraction
library(emmeans) # for model mean extraction
library(multcomp) # for pairwise comparison letter display
library(readxl) # for reading excel doc
library(janitor)
library(lmerTest)
library(dplyr)
```

# Introduction

Organic growers use a wide variety of fertilizers and materials to supply nutrients based on crop requirements. In the southeastern US, poultry litter (PL) is widely accessible and available for growers as a nutrient source, but growers must abide by the National Organic Program’s (NOP) “90-120-Day Rule”. The NOP specifies that manures must be applied 90 days before the harvest of food crops and 120 days before the harvest of food crops that have contact with the soil. The requirements of the 90-120-Day rule make it difficult to synchronize the mineralization of manures with crop demand, especially in terms of nitrogen. Applying the poultry litter to the cover crop, rather than at cash crop planting, allows the cover crop to uptake nutrients (nitrogen) from poultry litter with the potential to increase biomass and weed suppression, while reducing the risk of food-borne pathogens and the potential for nitrogen loss.

In conventional vegetable production, nitrogen availability in inorganic fertilizer is both plentiful and readily available for plant uptake. However, in organic production, farmers have of choice of costly certified organic fertilizers or other organic materials like compost, manures, and cover crops. Despite their accessibility and low costs, organic growers are hesitant to use manures such as poultry litter due to the risk of food-borne pathogens and regulations for using raw manure. The NOP states that manures must be applied ninety days before harvest for crops not in contact with the soil and 120 days before harvest for crops in contact with the soil. By applying broiler litter to cover crops, these plants can uptake the nutrients released from poultry litter, and as the cover crops decompose, release those nutrients for the use by cash crops without the risk of contamination by pathogens. Currently, the papers published in this field are all similar in the way of poultry application impacts on cover crops are studied and reported. Most of these studies look at biomass immediately before the termination of cover crops and only begin to measure soil inorganic nitrogen after cover crop termination. The study proposed here will vary from previous studies in testing the effects of poultry litter application on biomass and soil inorganic nitrogen from the seeding of the cover crop all the way through final harvest of the cash crop, instead of after the cover crop is terminated. Additionally, the study will examine how the quality of the cover crop affects subsequent nutrient availability and rate of release. The available literature also tends to focus on above ground biomass only and does not examine the effects on root biomass. The proposed project will also examine the role that root biomass and root nutrient content play in nutrient release and mineralization. In measuring these additional parameters and taking a more in-depth approach to previously conducted research, a better understanding of the uptake, release, and availability of nutrients can be formed.

The objective of this study is to determine the effects of poultry litter application to cover crops on cover crop biomass, weed suppression, soil inorganic nitrogen, and nutrient availability for the subsequent cash crop. To analyze the cover crop uptake of nutrients from poultry litter and subsequent release of nutrients from cover crop mineralization, the previously mentioned parameters will be measured before application of poultry litter then on days 4, 7, 14, and every two weeks until cover crop termination. The same sampling approach will be used for kale cycle from transplant to final harvest. Additionally, kale plant heights, chlorophyll measurement, and yield will be collected and analyzed.

Using the collected data through rigorous sampling dates, the proposed project could provide an in-depth understanding of nutrient uptake and release from cover crops and its impact on the following organic vegetable system.

# Hypothesis and Objectives

### Hypothesis

Poultry litter application will increase cover crop biomass, weed suppression, soil nitrogen concentrations, and nutrient availability for successive organic vegetable systems.

### Objectives

a.  Determine the effect of poultry litter application to cover crops on biomass, weed suppression, soil nitrogen concentrations, and nutrient availability.
b.  Determine the effect of poultry litter application on subsequent nitrogen availability released from cover crop.
c.  Determine the effect of poultry litter applications on cover crop quality as defined by cover crop composition of lignin, cellulose, and C:N ratio.

# Material and Methods

The two-year field study will take place on organic certified land at the University of Georgia’s Durham Horticulture farm in Watkinsville, GA. The farm has soils classified as a Cecil sandy loam by the United States Department of Agriculture (USDA). The plot will be 10 x 20 ft in a randomized complete block design. Four treatments of poultry litter will be applied at 0 (control treatment),1,2, and 4 tons per acre with four replication per treatment making a total of sixteen blocks. Organic oats will be seeded at a rate of 110 lbs per acre. At the end of the cover crop cycle, the oats will be flail-mowed and incorporated into the soil. Three weeks after incorporation of oats, the cash crop, kale (Brassica oleracea ‘Winterbor’) will be planted with 12” spacing in-row and 24” spacing between row with four rows per treatment (each treatment having fourty-eight plants total). Kale will be the subsequent cash crop as it is a staple crop for organic growers in the southeast and will be able to clearly reveal any differences in nitrogen accumulations between treatments when measuring various parameters.

To determine the effect of poultry litter on biomass of the cover crop, samples will be taken four times through the cover crop cycle using a 1 x 1 ft quadrat to harvest biomass at soil level, which will be dried at 65°C. Once dried, samples will be weighed, and their dry weight will be recorded. Additionally, light interception of the cover crops will be measured using the LP-80 (METER Group). Soil samples will be taken prior to poultry litter application, at 4,7, and 14 days after application then every two weeks until cover crop termination to determine soil inorganic nitrogen. The soil samples will be extracted in 1M KCl, then amounts of soil inorganic nitrogen will be calculated using a Timberline Instruments Ammonia Analyzer. Soil samples taken prior to poultry litter application will be analyzed for routine nutrient analysis, total nitrogen and carbon, and organic matter. At cover crop termination, biomass sample will be analyzed for total nitrogen, total carbon, carbohydrates, cellulose, and lignin using the NIRS DS5200 (Foss). Mineralization of these cover crops will be determined using the UGA Cover Crop Nitrogen Availability Calculator. Kale will be measured for nutrient uptake through soil samples taken at planting, and at 4,7, and 14 days after planting then every two weeks until final harvest. Leaf nitrate concentrations will be determined by harvesting two mature leaves and two new leaves per treatment block. Nitrate concentrations will be quantified using the same method used to determine soil nitrogen concentrations. To determine nitrogen availability for subsequent vegetable system, kale will be harvested four times throughout the study to and measured for nitrate, quality, dry weight, and overall yield.

# Study design

This study design is a randomized complete block design. There are 4 poultry litter application treatments with 4 replications randomized in 4 blocks (16 plots total). Randomization was conducted by take the 16 plot IDs and putting them into a web-generated randomized plot.

## Statistical analysis

Data will be analyzed using a type 3, analysis of variance (ANOVA) model in the program R (R Core Team). The independent (treatment) variable in the proposed will be the rate of poultry litter application, while dependent variables will include soil nitrogen, biomass, plant nitrate, plant height, and chlorophyll. When ANOVA indicated significance, Tukey's HSD was used to separate means at p \< 0.05. The models were run using the lmer function from the stats package in R. We checked linear model assumptions by extracting residuals and plotting on a QQ plot to look for linearity between our sample quantile and the theoretical. We checked for within group errors by plotting studentized residuals versus fitted ones, looking for outliers above 3 or below -3, while looking for possible patterns - where we did notice a fan shaped structure within our data. Which wasn’t something we had encountered or new how to fit our model to compensate for. Then with out studentized residuals we fitted to a QQ plot again to look for linearity between our sample quantile and the theoretical, which can back okay, with a linearity to the data without too much cause for concern. Finally we used geom_density function to look at the studentized residuals and how they lay between points of 3 and -3, to which we saw the majority of points fell near 0, with a clear bell-shaped pattern pattern. Overall when checking for normality or model seems to pass - besides the fan shape we noticed but that is outside the bounds of our course. For our work in R we used the following packages: tidyverse (for wrangling and plotting), car (for the Anova function), lme4 (to run our statistical model), broom (for model residual extractions), emmeans (for model means extraction), multicomp (for pairwise letters comparisons), readxl (to read and bring in our excel data), janitor (to wrangle and clean our excel data we import), LmerTest (to display the p-values next to the mdoel once run), and dyplr (for overall functionality)

##Bringing in Data 

Here we are bringing in our data set with the read_excel funtion through the readxl package which we configured and edited to run in R

```{r}
pulliam_proj <- read_excel("../data/project_raw_data.xlsx")

pulliam_proj
```

#EDA tables 

Here we are printing the summary and glimpsing at our data to get an understanding of how are variables are being interpreted and looking at the general stats so far with the raw data

```{r}
summary(pulliam_proj)

glimpse(pulliam_proj)
```

##Wrangling 

Here we are doing some light wrangling of the data, cleaning names to make coding easier down the line and primarily changing variables form being looked at as numerical to being viewed as factors, as they play important role in how are data is interpreted

```{r}
pulliam_proj_clean <- clean_names(pulliam_proj)

pulliam_proj_w <- pulliam_proj_clean %>%
  mutate(rep = factor(rep),
         date = factor(date),
         day = factor(day),
         trt = factor(trt),
         plot = factor(plot)
         )# %>%
 # mutate(trtname = paste0(,"+",))


#interested in mean effects so we changed variables to be viewed as factors 

pulliam_proj_w
```

##Sumamry 

Here we are looking at the summary again to ensure no mistakes were made

```{r}
summary(pulliam_proj_w)
```

# EDA Plots

Here we are trying to layout the data and look at possible presentations, sectioning out our main variables of trt and day to get an understanding on how it will impact our dependednt variable of total inorganic nitrogen (mg_kg_tin)

```{r fig.width=14, fig.height=6}
#sampling_dates <- c("Day 0", "Day 5", "Day 7", "Day 14", "Day 28", "Day 42", "Day 56", "Day 70","Day 93", "Day 119", "Day 4inc", "Day 14inc", "Day 6pk", "Day 14pk", "Day 30pk", "Day 47pk","Day 63pk", "Day 78pk", "Day 86pk")

#pulliam_proj_w$day <- factor(pulliam_proj_w$day, levels = original_day_order)

ggplot(pulliam_proj_w, aes(x = trt, 
                    y = mg_kg_tin,
                    color = trt)) +
  geom_boxplot() +
  geom_jitter() +
  theme(legend.position = "right") +
  
  
  facet_wrap(~ day)

# or could use facet_grid( ~ day) for one big graphic

ggplot(pulliam_proj_w, aes(x = trt, 
                    y = mg_kg_tin,
                    color = trt)) +
  geom_boxplot() +
  geom_jitter() +
  theme(legend.position = "right") +
  
  
  facet_grid(~ day)


```

# Statistical Model

Here we are creating our statistical model through the lme4 package using the lmer function, where we assume rep as randome and are attempting to exlain mg_kg_tin through the varaibles of trt and day

```{r}
options(contrasts = c("contr.sum", "contr.poly"))

pulliam_proj_mod <- lmer(mg_kg_tin ~ trt * day +
                       (1|rep),
                     data = pulliam_proj_w
                       )

pulliam_proj_mod
```

#Anova 
Here we are running a type 3 ANOVA model to look at significance

```{r}
Anova(pulliam_proj_mod, type=3)
```

#Linear Model Assumptions 
## Extracting residuals 

Here we are extracting residuals to begin chekcing normality

```{r}
library(broom.mixed)
pulliam_proj_mix_resid <- augment(pulliam_proj_mod) %>%
  mutate(.studresid=rstudent(pulliam_proj_mod))

pulliam_proj_mix_resid
```

##Random Effect are iid 

Here we are looking at radnom effect

```{r}
randeff_rep <- ranef(pulliam_proj_mod) [[1]]

randeff_rep
```

##QQPlot

Here we create a QQ plot to check the linear model assumptions fo the random effect

```{r}
ggplot(randeff_rep, aes(sample=`(Intercept)`))+
  stat_qq(  shape = 21,
            fill = "purple", 
            size = 3,
            alpha = .7
  )+
  stat_qq_line()+
  labs(x = "Theoretical quantile",
       y = "Sample quantile")+
  theme_bw()
```

##Within Group Errors are iid 

Here we plot residuals, studentized and fitted to look for outliers and normality issues (if we see a pattern)

```{r}
ggplot(pulliam_proj_mix_resid, aes(x=.fitted, y=.studresid))+
  geom_hline(yintercept = 0, color="red")+
  geom_point(shape = 21,
             fill = "purple", 
             size = 3,
             alpha = .7)+
  geom_smooth()+
  geom_hline(yintercept = c(-3,3), color = "red")+
  theme_bw()
```

##QQplot 

Here we check linear normality using the studentized residuals, keeping an eye out for unique patterns

```{r}
ggplot(pulliam_proj_mix_resid, aes(sample=.studresid))+
  stat_qq(  shape = 21,
            fill = "purple", 
            size = 3,
            alpha = .7
  )+
  stat_qq_line()+
  labs(x = "Theoretical quantile",
       y = "Sample quantile")+
  theme_bw()
```

##Geom Density Plot 

Here we look to ensure normality through the bell-shaped curve of the studentized resuals plotted

```{r}
ggplot(pulliam_proj_mix_resid, aes(x=.studresid))+
  geom_density(color = "black",
               fill = "purple",
               alpha = .7)+
  scale_x_continuous(breaks = c(-3,0,3), limits = c(-3,3))+
  theme_bw()
```

#Model Means 

Here we looking at model means bewteen individual treatments within a certain day

```{r}
pulliam_proj_means_all <- emmeans(pulliam_proj_mod, ~trt|day)

pulliam_proj_means_all
```

# Pairwise comparisons

Here we create our pairwise comparison and establish our letters of significance

```{r}
pulliam_proj_cld_all <- cld(pulliam_proj_means_all, 
                   reversed=T, 
                   adjust="none",
               Letters=letters)

pulliam_proj_cld_all
```

##Light Wrangling 

Here we are doing some light wrangling before moving on to our final plotting, but only needed if we decide to present the data in a way displaying "trt + day" together

```{r}
pulliam_proj_cld_selected <- pulliam_proj_cld_all %>%
  as.data.frame() %>%
  mutate(letter=trimws(.group)) %>%
  mutate(trtname = paste0(trt,"+", day))

pulliam_proj_cld_selected
```

#Final Plot 

Here we bring in our initial wrangled data to create boxplots with error bars, we set our dependent variable of mg_kg_tin to the y axis and our independent variables of day and treatment to the y axis, displaying both through the use of the facet_grid function. Our letters our created trough our emmean values we created a few steps above, then we save it in a size that makes it legible.

```{r fig.width=30, fig.height=8}
pulliam_proj_w1 <- pulliam_proj_w %>%
mutate(trtname = paste0(trt,"+", day))


ggplot(mapping = aes(fill = trtname))+
  # Raw data and boxplots  
  geom_boxplot(data = pulliam_proj_w1,
               aes(x = trt, y = mg_kg_tin),
               alpha = .8) +
  geom_jitter(data = pulliam_proj_w1,
               aes(x = trt, y = mg_kg_tin),
              shape = 21,
              size = 3,
              alpha = .6) +
  # Adding letters and titles
  geom_label(data = pulliam_proj_cld_selected,
            aes(x = trt, y = emmean, label = letter),
            fill = "white") +
  labs(x = "Treatments Grouped By Day",
       y = "Mg/Kg of Total Inorganic Nitrogen") +
  scale_fill_viridis_d() +
  theme(panel.grid.major.y = element_blank(),
        panel.grid.minor.y = element_blank(),
        legend.position = "none") +
  
facet_grid(~ day)

ggsave("../output/pulliam_project_final_plot.png",
       width = 30,
       height = 8)

```

# Results

Our graph shows no significance in soil total inorganic nitrogen on day 0 before poultry litter application. After day 0, there are significant differences in soil inorganic nitrogen from days 5-56 which makes sense. It is typical for poultry litter and other organic fertilizers to take a few weeks/month for microbes to bring their organic compounds into inorganic plant available forms which is what we are seeing in days 5 through 56. The lack of significant results from days 70-145 could be expected as cover crops are uptaking the inorganic nitrogen. On days 152 and days 185 there is a peak in soil inorganic nitrogen which is likely due to the mineralization of the incorporated cover crop in addition to the small pre-plant application of an organic fertilizer to supply the cash crop (kale) with enough N for the season. After this peak, there is another period of time with stable and non-significant soil inorganic nitrogen from days 185 to 224 which is likely due to the kale uptake of inorganic nitrogen. While treatment 4 (highest application of poultry litter) has the highest soil inorganic nitrogen as expected, we still need to analyze and visualize other nutrients/variables to examine the effects of this level of poultry litter applications.

The process and visualization of this data allows us to see how different rates of poultry litter provide different amounts of soil inorganic nitrogen at different times. Being able to visualize this allows for researcher and organic growers to make more informed applications of organic fertilizers to optimize and synchronize soil nutrients with plant uptake. Synchronization is a complex problem for organic growers as all organic fertilizers will mineralize and release their nutrients different depending on environmental and soil conditions. Being able to visualize this process can help minimize nutrient losses and increase fertilizer efficacy/yields. Further analysis will be conducted to see if poultry application to cover crops is suitable option for organic growers to avoid food pathogen risk while using low-cost fertilizers such as poultry litter.
