---
title: "Final Project"
author: "Hirsch-Santagata, Pulliam"
format: html
editor: visual
---

#Loading packages

```{r}
library(tidyverse) # for data wrangling and plotting
library(car) # for Anova function
library(lme4)
library(broom) # for model residuals extraction
library(emmeans) # for model mean extraction
library(multcomp) # for pairwise comparison letter display
library(readxl) # for reading excel doc
library(janitor)
library(lmerTest)
library(dplyr)
```

# Introduction

Organic growers use a wide variety of fertilizers and materials to supply nutrients based on crop requirements. In the southeastern US, poultry litter (PL) is widely accessible and available for growers as a nutrient source, but growers must abide by the National Organic Program’s (NOP) “90-120-Day Rule”. The NOP specifies that manures must be applied 90 days before the harvest of food crops and 120 days before the harvest of food crops that have contact with the soil. The requirements of the 90-120-Day rule make it difficult to synchronize the mineralization of manures with crop demand, especially in terms of nitrogen. Applying the poultry litter to the cover crop, rather than at cash crop planting, allows the cover crop to uptake nutrients (nitrogen) from poultry litter with the potential to increase biomass and weed suppression, while reducing the risk of food-borne pathogens and the potential for nitrogen loss.

In conventional vegetable production, nitrogen availability in inorganic fertilizer is both plentiful and readily available for plant uptake. However, in organic production, farmers have of choice of costly certified organic fertilizers or other organic materials like compost, manures, and cover crops. Despite their accessibility and low costs, organic growers are hesitant to use manures such as poultry litter due to the risk of food-borne pathogens and regulations for using raw manure. The NOP states that manures must be applied ninety days before harvest for crops not in contact with the soil and 120 days before harvest for crops in contact with the soil. By applying broiler litter to cover crops, these plants can uptake the nutrients released from poultry litter, and as the cover crops decompose, release those nutrients for the use by cash crops without the risk of contamination by pathogens. Currently, the papers published in this field are all similar in the way of poultry application impacts on cover crops are studied and reported. Most of these studies look at biomass immediately before the termination of cover crops and only begin to measure soil inorganic nitrogen after cover crop termination. The study proposed here will vary from previous studies in testing the effects of poultry litter application on biomass and soil inorganic nitrogen from the seeding of the cover crop all the way through final harvest of the cash crop, instead of after the cover crop is terminated. Additionally, the study will examine how the quality of the cover crop affects subsequent nutrient availability and rate of release. The available literature also tends to focus on above ground biomass only and does not examine the effects on root biomass. The proposed project will also examine the role that root biomass and root nutrient content play in nutrient release and mineralization. In measuring these additional parameters and taking a more in-depth approach to previously conducted research, a better understanding of the uptake, release, and availability of nutrients can be formed.

The objective of this study is to determine the effects of poultry litter application to cover crops on cover crop biomass, weed suppression, soil inorganic nitrogen, and nutrient availability for the subsequent cash crop. To analyze the cover crop uptake of nutrients from poultry litter and subsequent release of nutrients from cover crop mineralization, the previously mentioned parameters will be measured before application of poultry litter then on days 4, 7, 14, and every two weeks until cover crop termination. The same sampling approach will be used for kale cycle from transplant to final harvest. Additionally, kale plant heights, chlorophyll measurement, and yield will be collected and analyzed.

Using the collected data through rigorous sampling dates, the proposed project could provide an in-depth understanding of nutrient uptake and release from cover crops and its impact on the following organic vegetable system.

# Hypothesis and objectives

### Hypothesis

Poultry litter application will increase cover crop biomass, weed suppression, soil nitrogen concentrations, and nutrient availability for successive organic vegetable systems.

### Objectives           

a.  Determine the effect of poultry litter application to cover crops on biomass, weed suppression, soil nitrogen concentrations, and nutrient availability.
b.  Determine the effect of poultry litter application on subsequent nitrogen availability released from cover crop.
c.  Determine the effect of poultry litter applications on cover crop quality as defined by cover crop composition of lignin, cellulose, and C:N ratio.

# Material and Methods

The two-year field study will take place on organic certified land at the University of Georgia’s Durham Horticulture farm in Watkinsville, GA. The farm has soils classified as a Cecil sandy loam by the United States Department of Agriculture (USDA). The plot will be 10 x 20 ft in a randomized complete block design. Four treatments of poultry litter will be applied at 0 (control treatment),1,2, and 4 tons per acre with four replication per treatment making a total of sixteen blocks. Organic oats will be seeded at a rate of 110 lbs per acre. At the end of the cover crop cycle, the oats will be flail-mowed and incorporated into the soil. Three weeks after incorporation of oats, the cash crop, kale (Brassica oleracea ‘Winterbor’) will be planted with 12” spacing in-row and 24” spacing between row with four rows per treatment (each treatment having fourty-eight plants total). Kale will be the subsequent cash crop as it is a staple crop for organic growers in the southeast and will be able to clearly reveal any differences in nitrogen accumulations between treatments when measuring various parameters.

To determine the effect of poultry litter on biomass of the cover crop, samples will be taken four times through the cover crop cycle using a 1 x 1 ft quadrat to harvest biomass at soil level, which will be dried at 65°C. Once dried, samples will be weighed, and their dry weight will be recorded. Additionally, light interception of the cover crops will be measured using the LP-80 (METER Group). Soil samples will be taken prior to poultry litter application, at 4,7, and 14 days after application then every two weeks until cover crop termination to determine soil inorganic nitrogen. The soil samples will be extracted in 1M KCl, then amounts of soil inorganic nitrogen will be calculated using a Timberline Instruments Ammonia Analyzer. Soil samples taken prior to poultry litter application will be analyzed for routine nutrient analysis, total nitrogen and carbon, and organic matter. At cover crop termination, biomass sample will be analyzed for total nitrogen, total carbon, carbohydrates, cellulose, and lignin using the NIRS DS5200 (Foss). Mineralization of these cover crops will be determined using the UGA Cover Crop Nitrogen Availability Calculator. Kale will be measured for nutrient uptake through soil samples taken at planting, and at 4,7, and 14 days after planting then every two weeks until final harvest. Leaf nitrate concentrations will be determined by harvesting two mature leaves and two new leaves per treatment block. Nitrate concentrations will be quantified using the same method used to determine soil nitrogen concentrations. To determine nitrogen availability for subsequent vegetable system, kale will be harvested four times throughout the study to and measured for nitrate, quality, dry weight, and overall yield.

Data will be analyzed using a one way analysis of variance (ANOVA) model in the program R (R Core Team). The independent (treatment) variable in the proposed will be the rate of poultry litter application, while dependent variables will include soil nitrogen, biomass, plant nitrate, plant height, and chlorophyll. When ANOVA indicatedd significance, Tukey's HSD was used o superate means at p < 0.05. The models were run using the lm function from the stats package in R.

# Study design

This study design is a randomized complete block design. There are 4 poultry litter application treatments with 4 replications randomized in 4 blocks (16 plots total). Randomization was conducted by take the 16 plot IDs and putting them into a web-generated randomized plot.



#Bringing in Data

```{r}
pulliam_proj <- read_excel("../data/project_raw_data.xlsx")

pulliam_proj
```

#EDA tables

```{r}
summary(pulliam_proj)

glimpse(pulliam_proj)
```

#Wrangling

```{r}
pulliam_proj_clean <- clean_names(pulliam_proj)

pulliam_proj_w <- pulliam_proj_clean %>%
  mutate(rep = factor(rep),
         date = factor(date),
         day = factor(day),
         trt = factor(trt),
         plot = factor(plot)
         )# %>%
 # mutate(trtname = paste0(,"+",))


#interested in mean effects so we changed variables to be viewed as factors 

pulliam_proj_w
```

#sumamry

```{r}
summary(pulliam_proj_w)
```

# EDA Plots

```{r fig.width=14, fig.height=6}
#sampling_dates <- c("Day 0", "Day 5", "Day 7", "Day 14", "Day 28", "Day 42", "Day 56", "Day 70","Day 93", "Day 119", "Day 4inc", "Day 14inc", "Day 6pk", "Day 14pk", "Day 30pk", "Day 47pk","Day 63pk", "Day 78pk", "Day 86pk")

#pulliam_proj_w$day <- factor(pulliam_proj_w$day, levels = original_day_order)

ggplot(pulliam_proj_w, aes(x = trt, 
                    y = mg_kg_tin,
                    color = trt)) +
  geom_boxplot() +
  geom_jitter() +
  theme(legend.position = "right") +
  
  
  facet_wrap(~ day)

# or could use facet_grid( ~ day) for one big graphic

ggplot(pulliam_proj_w, aes(x = trt, 
                    y = mg_kg_tin,
                    color = trt)) +
  geom_boxplot() +
  geom_jitter() +
  theme(legend.position = "right") +
  
  
  facet_grid(~ day)


```

#Faceting longer by day? (or sampling date)

```{r}
#pulliam_wide <- pulliam_proj_w %>%
  #unite("day_trt", day, trt, sep = "_") %>%
 # pivot_wider(names_from = day_trt, values_from = mg_kg_tin,
           #   values_fill = 0) ## this last bit is to ake NA to 0 

#pulliam_wide


```

# Statistical Model

```{r}
options(contrasts = c("contr.sum", "contr.poly"))

pulliam_proj_mod <- lmer(mg_kg_tin ~ trt * day +
                       (1|rep),
                     data = pulliam_proj_w
                       )

pulliam_proj_mod
```

#Anova

```{r}
Anova(pulliam_proj_mod, type=3)
```

#Linear Model Assumptions \## extracting residuals

```{r}
library(broom.mixed)
pulliam_proj_mix_resid <- augment(pulliam_proj_mod) %>%
  mutate(.studresid=rstudent(pulliam_proj_mod))

pulliam_proj_mix_resid
```

#Random effect are iid

```{r}
randeff_rep <- ranef(pulliam_proj_mod) [[1]]

randeff_rep
```

##QQPLot

```{r}
ggplot(randeff_rep, aes(sample=`(Intercept)`))+
  stat_qq(  shape = 21,
            fill = "purple", 
            size = 3,
            alpha = .7
  )+
  stat_qq_line()+
  labs(x = "Theoretical quantile",
       y = "Sample quantile")+
  theme_bw()
```

##within group errors are iid

```{r}
ggplot(pulliam_proj_mix_resid, aes(x=.fitted, y=.studresid))+
  geom_hline(yintercept = 0, color="red")+
  geom_point(shape = 21,
             fill = "purple", 
             size = 3,
             alpha = .7)+
  geom_smooth()+
  geom_hline(yintercept = c(-3,3), color = "red")+
  theme_bw()
```

```{r}
ggplot(pulliam_proj_mix_resid, aes(sample=.studresid))+
  stat_qq(  shape = 21,
            fill = "purple", 
            size = 3,
            alpha = .7
  )+
  stat_qq_line()+
  labs(x = "Theoretical quantile",
       y = "Sample quantile")+
  theme_bw()
```

```{r}
ggplot(pulliam_proj_mix_resid, aes(x=.studresid))+
  geom_density(color = "black",
               fill = "purple",
               alpha = .7)+
  scale_x_continuous(breaks = c(-3,0,3), limits = c(-3,3))+
  theme_bw()
```

#Model Means

```{r}
pulliam_proj_means_all <- emmeans(pulliam_proj_mod, ~trt:day)

pulliam_proj_means_all
```
